---
title: "07-30-2014_pedestrians"
author: "Gabriel Florit"
output:
  html_document:
    self_contained: false
---

```{r, echo=FALSE, results='hide', warning=FALSE, message=FALSE}
setwd("~/Documents/dev/07-30-2014_pedestrians/data")

# Multiple plot function
#
# ggplot objects can be passed in ..., or to plotlist (as a list of ggplot objects)
# - cols:   Number of columns in layout
# - layout: A matrix specifying the layout. If present, 'cols' is ignored.
#
# If the layout is something like matrix(c(1,2,3,3), nrow=2, byrow=TRUE),
# then plot 1 will go in the upper left, 2 will go in the upper right, and
# 3 will go all the way across the bottom.
#
multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
  require(grid)

  # Make a list from the ... arguments and plotlist
  plots <- c(list(...), plotlist)

  numPlots = length(plots)

  # If layout is NULL, then use 'cols' to determine layout
  if (is.null(layout)) {
    # Make the panel
    # ncol: Number of columns of plots
    # nrow: Number of rows needed, calculated from # of cols
    layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
                    ncol = cols, nrow = ceiling(numPlots/cols))
  }

 if (numPlots==1) {
    print(plots[[1]])

  } else {
    # Set up the page
    grid.newpage()
    pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))

    # Make each plot, in the correct location
    for (i in 1:numPlots) {
      # Get the i,j matrix positions of the regions that contain this subplot
      matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))

      print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
                                      layout.pos.col = matchidx$col))
    }
  }
}

library(stringr)
library(lubridate)
library(RColorBrewer)
library(rgdal)
library(rgeos)
library(tidyr)
library(dplyr)
library(ggmap)
library(ggplot2)

# read shapefiles
crashes.shp <- readOGR('downloaded/pedestriancrashes/shp','pedestriancrashes',  p4s='+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs')
tracts.shp <- readOGR('downloaded','cb_2013_25_tract_500k',p4s='+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs')
state.shp <- readOGR('downloaded','cb_2013_ma_state_500k', p4s='+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs')

# add tract information to each crash
# standardize dates
# reject bad dates, e.g. 2041
crashes <- over(crashes.shp, tracts.shp) %>%
  cbind(crashes.shp@data) %>%
  mutate(DATE = mdy(Crash.Date)) %>%
  filter(year(today()) >= year(DATE))

# calculate crashes per tract
crashesPerTract <- crashes %>%
  group_by(GEOID) %>%
  summarise(crashes = n()) %>%
  arrange(desc(crashes))

tracts <- tracts.shp@data %>%
  left_join(crashesPerTract,by='GEOID') %>%
  arrange(desc(crashes))
  
# fortify for mapping purposes
tracts.fortified <- fortify(tracts.shp,region='GEOID')
state.fortified <- fortify(state.shp,region='GEOID')

# add more data to tracts.shp
#tracts.shp@data <- 

# add more data to tracts.fortified
#tracts.fortified <- tracts.fortified %>%
#  mutate(GEOID=id) %>%
#  left_join(crashesPerTract,by='GEOID')

#%>%
#  
# add tract information to crashes
#crashes.shp@data <- over(crashes.shp, tracts.shp) %>%
#  cbind(crashes.shp@data) %>%
#  mutate(DATE = mdy(Crash.Date))

#  group_by(GEOID) %>%
#  summarise(crashes = n()) %>%
#  arrange(desc(crashes))

# calculate tract centroids
#tracts.shp@data <- cbind(
#  gCentroid(tracts.shp,byid=T)@coords,
#  tracts.shp@data) %>%
#  left_join(crashesInTracts,by='GEOID')
```

Let's explore the question: are pedestrian crashes increasing?

```{r, echo=FALSE, fig.width=12, fig.height=7}

#yearsRange <- crashes %>%
#  mutate(
#    min = min(year(DATE)),
#    max = max(year(DATE))
#  ) %>%
#  head(1)

#data <- tracts %>%
#  filter(!is.na(crashes))

#ggplot() +
#  geom_map(data=data,aes(map_id=GEOID,fill=crashes),map=tracts.fortified) +
#  geom_path(data=state.fortified,aes(long,lat,group=group),colour='black',size=0.125) +
#  coord_map() +
#  scale_fill_gradientn(name='Crashes',colours=brewer.pal(9,'Reds')) +
#  ggtitle(str_c('Pedestrian crashes by census tract, ',yearsRange[1,]$min,'-',yearsRange[1,]$max)) +
#  theme_minimal() +
#  theme(
#    panel.grid = element_blank(),
#    axis.ticks = element_blank(),
#    axis.text = element_blank(),
#    axis.title = element_blank()
#  )
```

***

Let's look at the ten worst census tracts.

```{r, echo=FALSE, fig.width=12, fig.height=24, warning=FALSE, message=FALSE}

#data <- tracts.shp@data %>%
#  arrange(desc(crashes))

#crashPoints <- as.data.frame(crashes.shp@coords) %>%
#  mutate(long=coords.x1,lat=coords.x2) %>% select(-coords.x1,-coords.x2) %>%
#  group_by(long,lat) %>%
#  summarise(count=n()) %>%
#  arrange(desc(count))

map1 <- ggmap(get_map(location=c(lon=data[1,]$x,lat=data[1,]$y),zoom=15)) +
  geom_polygon(data=tracts, aes(x=long, y=lat, group=group, fill=crashes), colour='NA', alpha=0.65)+
  geom_point(data=crashPoints,aes(x=long,y=lat,size=count)) +
  scale_size_area(max_size=40) +
  scale_fill_gradientn(name='Pedestrian crashes',colours=brewer.pal(9,'Reds')) +
  theme_nothing()

map2 <- ggmap(get_map(location=c(lon=data[2,]$x,lat=data[2,]$y),zoom=15)) +
  geom_polygon(data=tracts, aes(x=long, y=lat, group=group, fill=crashes), colour='NA', alpha=0.65)+
  geom_point(data=crashPoints,aes(x=long,y=lat,size=count)) +
  scale_size_area(max_size=40) +
  scale_fill_gradientn(name='Pedestrian crashes',colours=brewer.pal(9,'Reds')) +
  theme_nothing()

map3 <- ggmap(get_map(location=c(lon=data[3,]$x,lat=data[3,]$y),zoom=15)) +
  geom_polygon(data=tracts, aes(x=long, y=lat, group=group, fill=crashes), colour='NA', alpha=0.65)+
  geom_point(data=crashPoints,aes(x=long,y=lat,size=count)) +
  scale_size_area(max_size=40) +
  scale_fill_gradientn(name='Pedestrian crashes',colours=brewer.pal(9,'Reds')) +
  theme_nothing()

map4 <- ggmap(get_map(location=c(lon=data[4,]$x,lat=data[4,]$y),zoom=15)) +
  geom_polygon(data=tracts, aes(x=long, y=lat, group=group, fill=crashes), colour='NA', alpha=0.65)+
  geom_point(data=crashPoints,aes(x=long,y=lat,size=count)) +
  scale_size_area(max_size=40) +
  scale_fill_gradientn(name='Pedestrian crashes',colours=brewer.pal(9,'Reds')) +
  theme_nothing()

map5 <- ggmap(get_map(location=c(lon=data[5,]$x,lat=data[5,]$y),zoom=15)) +
  geom_polygon(data=tracts, aes(x=long, y=lat, group=group, fill=crashes), colour='NA', alpha=0.65)+
  geom_point(data=crashPoints,aes(x=long,y=lat,size=count)) +
  scale_size_area(max_size=40) +
  scale_fill_gradientn(name='Pedestrian crashes',colours=brewer.pal(9,'Reds')) +
  theme_nothing()

map6 <- ggmap(get_map(location=c(lon=data[6,]$x,lat=data[6,]$y),zoom=15)) +
  geom_polygon(data=tracts, aes(x=long, y=lat, group=group, fill=crashes), colour='NA', alpha=0.65)+
  geom_point(data=crashPoints,aes(x=long,y=lat,size=count)) +
  scale_size_area(max_size=40) +
  scale_fill_gradientn(name='Pedestrian crashes',colours=brewer.pal(9,'Reds')) +
  theme_nothing()

map7 <- ggmap(get_map(location=c(lon=data[7,]$x,lat=data[7,]$y),zoom=15)) +
  geom_polygon(data=tracts, aes(x=long, y=lat, group=group, fill=crashes), colour='NA', alpha=0.65)+
  geom_point(data=crashPoints,aes(x=long,y=lat,size=count)) +
  scale_size_area(max_size=40) +
  scale_fill_gradientn(name='Pedestrian crashes',colours=brewer.pal(9,'Reds')) +
  theme_nothing()

map8 <- ggmap(get_map(location=c(lon=data[8,]$x,lat=data[8,]$y),zoom=15)) +
  geom_polygon(data=tracts, aes(x=long, y=lat, group=group, fill=crashes), colour='NA', alpha=0.65)+
  geom_point(data=crashPoints,aes(x=long,y=lat,size=count)) +
  scale_size_area(max_size=40) +
  scale_fill_gradientn(name='Pedestrian crashes',colours=brewer.pal(9,'Reds')) +
  theme_nothing()

#multiplot(map1, map2, map3, map4, map5, map6, map7, map8, cols=2)
```

No surprises here - crashes are happening in areas with high pedestrian traffic.

***

Are pedestrian crashes increasing over time?

```{r, echo=FALSE, fig.width=12, fig.height=7, warning=FALSE, message=FALSE}

#data <- crashes.shp@data %>%
#  filter(as.character(City.Town) == 'BOSTON')
  
  
#  select(DATE) %>%
#  mutate(DATE = floor_date(DATE, 'week')) %>%
#  filter(year(DATE) %in% c(2010,2011,2012)) %>%
#  group_by(DATE) %>%
#  summarise(count = n()) %>%
#  arrange(desc(DATE))

#ggplot(data, aes(DATE,count)) +
#  geom_line() +
#  geom_smooth() +
#  ggtitle('Pedestrian crashes, 2010-2012') +
#  xlab('week')
```

Maybe a little bit - but perhaps the state population is also increasing.

***

What about daily patterns?

```{r, echo=FALSE, fig.width=12, fig.height=7, warning=FALSE, message=FALSE}

data <- crashes.shp@data %>%
  mutate(HOUR = hour(parse_date_time(Crash.Time, 'I:M p'))) %>%
  select(Crash.Time, HOUR) %>%
  group_by(HOUR) %>%
  summarise(count = n())

#ggplot(data, aes(HOUR,count)) +
#  geom_line() +
#  geom_smooth() +
#  ggtitle('Pedestrian crashes per hour, 2010-2012') +
#  xlab('hour')
```

Crashes peak in the evening, when people are commuting home. No surprises here.

***

What about fatal crashes - are they increasing?

```{r, echo=FALSE, fig.width=12, fig.height=7, warning=FALSE, message=FALSE}

data <- crashes.shp@data %>%
  mutate(
    DATE = floor_date(DATE, 'week'),
    FATALITIES = as.numeric(as.character(Total.Fata))
  ) %>%
  filter(
    year(DATE) %in% c(2010,2011,2012),
    FATALITIES > 0
  ) %>%
  group_by(DATE) %>%
  summarise(count = sum(FATALITIES)) %>%
  arrange(desc(DATE))

#ggplot(data, aes(DATE,count)) +
#  geom_line() +
#  geom_smooth() +
#  ggtitle('Pedestrian fatalities, 2010-2012') +
#  xlab('week')
```

Perhaps - but it's tough to see when we group crashes by week.

***

Let's group crashes by year instead:

```{r, echo=FALSE, fig.width=12, fig.height=7, warning=FALSE, message=FALSE}

data <- crashes.shp@data %>%
  mutate(
    DATE = floor_date(DATE, 'year'),
    FATALITIES = as.numeric(as.character(Total.Fata))
  ) %>%
  filter(
    year(DATE) %in% c(2010,2011,2012),
    FATALITIES > 0
  ) %>%
  group_by(DATE) %>%
  summarise(count = sum(FATALITIES)) %>%
  arrange(desc(DATE))

#ggplot(data, aes(DATE,count)) +
#  geom_line() +
#  ggtitle('Pedestrian fatalities, 2010-2012') +
#  xlab('year')
```

We don't have enough data to see if fatalities are increasing.

***

What are vehicles doing prior to the crash?

```{r, echo=FALSE, fig.width=12, fig.height=7, warning=FALSE, message=FALSE}

data <- crashes.shp@data %>%
  group_by(Vehicle.Ac) %>%
  summarise(count = n()) %>%
  arrange(desc(count))

#data
```

No patterns here.

***

It would be great if we had more years than just 2010-2012. Fortunately for us, MassDOT's "Crash Portal" does just that. Let's look at all pedestrian crashes over time:

```{r, echo=FALSE, fig.width=12, fig.height=7, warning=FALSE, message=FALSE}

allcrashes <- read.csv('private/allcrashes.csv',strip.white=T,stringsAsFactors=F)

data <- allcrashes %>%
  mutate(DATE = floor_date(mdy(Crash.Date), 'year')) %>%
  group_by(DATE) %>%
  summarise(count = n()) %>%
  filter(year(DATE) < 2015 & year(DATE) >= 2002) %>%
  filter(year(DATE) %in% c(2010,2011,2012)) %>%
  arrange(DATE)

#ggplot(data, aes(DATE,count)) +
#  geom_line() +
#  ggtitle('Pedestrian crashes, 2002-2012') +
#  xlab('year')
```

Aha! Pedestrian crashes have increased dramatically since 2009.
