---
title: "07-30-2014_pedestrians"
author: "Gabriel Florit"
output:
  html_document:
    self_contained: false
---

```{r, echo=FALSE, results='hide', warning=FALSE, message=FALSE}
setwd("~/Documents/dev/07-30-2014_pedestrians/data")

# Multiple plot function
#
# ggplot objects can be passed in ..., or to plotlist (as a list of ggplot objects)
# - cols:   Number of columns in layout
# - layout: A matrix specifying the layout. If present, 'cols' is ignored.
#
# If the layout is something like matrix(c(1,2,3,3), nrow=2, byrow=TRUE),
# then plot 1 will go in the upper left, 2 will go in the upper right, and
# 3 will go all the way across the bottom.
#
multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
  require(grid)

  # Make a list from the ... arguments and plotlist
  plots <- c(list(...), plotlist)

  numPlots = length(plots)

  # If layout is NULL, then use 'cols' to determine layout
  if (is.null(layout)) {
    # Make the panel
    # ncol: Number of columns of plots
    # nrow: Number of rows needed, calculated from # of cols
    layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
                    ncol = cols, nrow = ceiling(numPlots/cols))
  }

 if (numPlots==1) {
    print(plots[[1]])

  } else {
    # Set up the page
    grid.newpage()
    pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))

    # Make each plot, in the correct location
    for (i in 1:numPlots) {
      # Get the i,j matrix positions of the regions that contain this subplot
      matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))

      print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
                                      layout.pos.col = matchidx$col))
    }
  }
}

library(zoo)
library(stringr)
library(lubridate)
library(RColorBrewer)
library(rgdal)
library(rgeos)
library(tidyr)
library(dplyr)
library(ggmap)
library(ggplot2)
```

Let's explore the question: are pedestrian crashes increasing? We'll be working with the following data sources:

- pedestrian crashes from MassDOT's [Crash Portal](http://services.massdot.state.ma.us/crashportal/CrashMapPage.aspx?Mode=Adhoc)
- town shapefile from [MassGIS](http://www.mass.gov/anf/research-and-tech/it-serv-and-support/application-serv/office-of-geographic-information-massgis/datalayers/towns.html)
- town population estimates [from US Census Bureau](https://www.census.gov/popest/data/datasets.html)

### Setup

```{r, results='hide', warning=FALSE, message=FALSE}

# read shapefiles
crashes.shp <- readOGR('downloaded/pedestriancrashes/shp','pedestriancrashes',  p4s='+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs')
towns.shp <- readOGR('downloaded','MA_TOWNS', p4s='+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs')

# standardize dates
# reject bad dates, e.g. 2041
crashes <- crashes.shp@data %>%
  mutate(DATE = mdy(Crash.Date)) %>%
  filter(year(today()) >= year(DATE))

# calculate crashes per year for each town
yearlyCrashesPerTown <- crashes %>%
  mutate(YEAR=year(DATE)) %>%
  group_by(YEAR,City.Town) %>%
  summarise(crashes = n())

# create towns dataframe
# add town boundaries
# add total crashes per town
towns <- towns.shp@data %>%
  cbind(fortify(gEnvelope(towns.shp,byid=T)) %>%
          mutate(id=as.numeric(id)) %>%
          group_by(id) %>%
          summarise(E=min(long),W=max(long),S=min(lat),N=max(lat)) %>%
          arrange(id)) %>%
  left_join(yearlyCrashesPerTown %>%
              group_by(City.Town) %>%
              summarise(crashes = sum(crashes)) %>%
              mutate(TOWN=City.Town) %>% select(-City.Town),by='TOWN')

# fortify for mapping purposes
towns.fortified <- fortify(towns.shp,region='TOWN')

# read population data
# remove ' Town' or ' City' suffixes from town names 
population <- read.csv('downloaded/towns_2000-2010s.csv',strip.white=T,stringsAsFactors=F) %>%
  gather(year,population,-NAME) %>%
  mutate(
    YEAR = as.numeric(gsub('POPESTIMATE', '', as.character(year))),
    place = gsub(' TOWN$| CITY$| TOWN CITY$', '', toupper(NAME))
  ) %>%
  select(-NAME,-year)
```

### Analysis

Are pedestrian crashes increasing?

```{r, fig.width=12, fig.height=7}

# group crashes by year
# get count
# only look at 2002 onwards
data <- crashes %>%
  mutate(DATE=floor_date(DATE,'year')) %>%
  group_by(DATE) %>%
  summarise(count = n()) %>%
  filter(year(DATE) >= 2002) %>%
  arrange(DATE)

ggplot(data,aes(DATE,count)) +
  geom_line() +
  ggtitle(str_c('Massachusetts pedestrian crashes, ',year(min(data$DATE)),'-',year(max(data$DATE)))) +
  xlab('year') +
  ylim(c(0,max(data$count)))
```

It seems so.

***

But maybe the population is increasing as well? What if we normalize by population?

```{r, fig.width=12, fig.height=7}

# group crashes by year
# get count
# only look at 2002 onwards
# add yearly state population
# calculate yearly pedestrian accident rate
data <- crashes %>%
  mutate(YEAR=year(DATE)) %>%
  group_by(YEAR) %>%
  summarise(count = n()) %>%
  filter(YEAR >= 2002) %>%
  mutate(place='MASSACHUSETTS') %>%
  left_join(population,by=c('YEAR','place')) %>%
  mutate(
    DATE=as.Date(as.yearmon(YEAR)),
    rate=count/population * 100
  ) %>%
  arrange(YEAR)

ggplot(data,aes(DATE,rate)) +
  geom_line() +
  ggtitle(str_c('Massachusetts pedestrian crashes per capita, ',year(min(data$DATE)),'-',year(max(data$DATE)))) +
  xlab('year') +
  ylab('crashes per 10,000') +
  ylim(c(0,max(data$rate)))
```

The state's population hasn't increased significantly, hence the rate looks pretty similar to the raw count.

***

Can we analyse this data to see which towns are working on their reporting?

```{r, fig.width=12, fig.height=7}

# get top towns by total crashes
data <- towns %>%
  arrange(desc(crashes)) %>%
  head(5) %>%
  select(City.Town=TOWN) %>%
  left_join(yearlyCrashesPerTown,by='City.Town') %>%
  mutate(DATE=as.Date(as.yearmon(YEAR))) %>%
  filter(year(DATE) >= 2002)

ggplot(data,aes(DATE,crashes,group=City.Town)) +
  geom_line(aes(color=City.Town=='BOSTON')) +
  geom_text(data=filter(data,YEAR==max(YEAR)),aes(DATE,crashes,label=City.Town),hjust=1,vjust=-0.5,size=3) +
  ggtitle(str_c('Towns with most pedestrian crashes, ',min(data$YEAR),'-',max(data$YEAR))) +
  theme(legend.position='none') +
  ylim(c(0,max(data$crashes)))
```

Boston's reporting inconsistency is striking.

***

What if we normalize by population?

```{r, fig.width=12, fig.height=7}

# get top towns by total crashes
# add yearly town population
data <- towns %>%
  arrange(desc(crashes)) %>%
  head(5) %>%
  select(City.Town=TOWN) %>%
  left_join(yearlyCrashesPerTown,by='City.Town') %>%
  mutate(
    DATE=as.Date(as.yearmon(YEAR)),
    place=as.character(City.Town)
  ) %>%
  select(-City.Town) %>%
  filter(year(DATE) >= 2002) %>%
  left_join(population,by=c('YEAR','place')) %>%
  mutate(rate=crashes/population * 10000)

ggplot(data,aes(DATE,rate,group=place)) +
  geom_line(aes(color=place=='BOSTON')) +
  geom_text(data=filter(data,YEAR==max(YEAR)),aes(DATE,rate,label=place),hjust=1,vjust=-0.5,size=3) +
  ggtitle(str_c('Pedestrian crashes per capita, ',min(data$YEAR),'-',max(data$YEAR))) +
  ylab('crashes per 10,000') +
  theme(legend.position='none') +
  ylim(c(0,max(data$rate)))
```

This graph shows that Boston's reported crash rate is quite low. We could probably make a case that Boston has a substantial amount of external foot traffic - people that don't live in the city but work in the city. If we were to include that number, Boston's crash rate would go down even lower.

***

What about other towns?

```{r, fig.width=12, fig.height=12}

# only look at towns with more than 50 crashes total
data <- towns %>%
  filter(crashes>50) %>%
  arrange(desc(crashes)) %>%
  select(City.Town=TOWN) %>%
  left_join(yearlyCrashesPerTown,by='City.Town') %>%
  mutate(DATE=as.Date(as.yearmon(YEAR))) %>%
  filter(year(DATE) >= 2002)

ggplot(data,aes(DATE,crashes,group=City.Town)) +
  geom_line() +
  ggtitle(str_c('Pedestrian crashes in towns with greater than 50 crashes in period ',min(data$YEAR),'-',max(data$YEAR),'. Y-scale is relative to each town.')) +
  facet_wrap(~City.Town,scales='free',ncol=8) +
  theme(
    axis.ticks=element_blank(),
    axis.text=element_blank(),
    panel.grid = element_blank()
  )
```

It seems that Barnstable and Springfield didn't report pedestrian accidents prior to 2010.

***

One final graph - let's look at the distribution of reported crashes per town over time.

```{r, fig.width=12, fig.height=7, warning=FALSE, message=FALSE}

data <- towns %>%
  arrange(desc(crashes)) %>%
  select(City.Town=TOWN) %>%
  left_join(yearlyCrashesPerTown,by='City.Town') %>%
  mutate(DATE=as.Date(as.yearmon(YEAR))) %>%
  filter(year(DATE) >= 2002)

ggplot(data,aes(crashes,group=YEAR)) +
  geom_histogram() +
  facet_wrap(~YEAR,nrow=3) +
  ggtitle(str_c('Distribution of pedestrian crashes per town, ',min(data$YEAR),'-',max(data$YEAR)))
```

That's not particularly useful.

```{r, fig.width=12, fig.height=7, warning=FALSE, message=FALSE}

data <- towns %>%
  arrange(desc(crashes)) %>%
  select(City.Town=TOWN) %>%
  left_join(yearlyCrashesPerTown,by='City.Town') %>%
  mutate(DATE=as.Date(as.yearmon(YEAR))) %>%
  filter(year(DATE) >= 2002)

ggplot(data,aes(DATE,crashes,group=DATE)) +
  geom_boxplot() +
  ggtitle(str_c('Distribution of pedestrian crashes per town, ',min(data$YEAR),'-',max(data$YEAR)))
```

Interesting, but not very revealing.

***

How many pedestrian accidents (involving injuries) in Boston?

```{r, fig.width=12, fig.height=7, warning=FALSE, message=FALSE, results='asis'}

# create a new column, SEVERITY, that creates three groups: fatal injury, non-fatal injury, other
data <- crashes %>%
  mutate(
    YEAR = year(DATE),
    SEVERITY = ifelse(grepl('fatal injury',tolower(as.character(Crash.Seve))), as.character(Crash.Seve), 'Other')
  ) %>%
  filter(
    City.Town == 'BOSTON',
    YEAR > 2009
  ) %>%
  group_by(YEAR,SEVERITY) %>%
  summarise(ACCIDENTS = n())

knitr::kable(data)
```